Type theory with single substitutions

### 2025-02-21 cpp/p.tex
minimisation

\title{Type theory in type theory with single substitutions}

@@

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck the other equations. SSC provides a simple, minimalistic
alternative to parallel substitution calculi or B-systems for defining
type theory. SSC relates to CwF as extensional combinatory calculus
relates to lambda calculus: there are more models of the former, but
the syntaxes are equivalent. If we have some additional type formers,
we show that SSC models are actually equivalent to CwF-based models in
a weak sense. Sections 2 and 3 of this paper were formalised in Agda.
\end{abstract}

@@

In this paper, we will answer this question by defining type theory as
a generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). GATs are multi-sorted algebraic
theories where later sorts can be indexed by previous sorts. An
example is categories where the sort of morphisms is double-indexed
over objects.

The GAT presentation of type theory is intrinsic
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} rather
than extrinsic
\cite{alti:phd93,abel2013normalization,DBLP:conf/cpp/AdjedjLMPP24}:
this means that we only consider well-formed, well-scoped and
well-typed abstract syntax trees, there are no meaningless terms; in
the GAT presentation, the conversion relation is expressed by
equations of the algebraic theory.

@@

Our SSC is in some sense too minimalistic: for example, the equation
$b[\p^+][\langle\q\rangle] = b$ is not \emph{derivable} in any model,
but it is \emph{admissible} in the syntax. (Here $b : \Tm\,(\Gamma\ext
A)\,B$ is a term which depends on a context $\Gamma$ and an extra
variable of type $A$; $b[\p^+] : \Tm\,(\Gamma\ext A\ext
A[\p])\,(B[\p^+])$ is a version where we weakened \emph{before} the
last variable; in $b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$
we substituted the last variable for the previous one.)  This is
analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In this paper, we show
that the \emph{syntaxes} of our SSC and the corresponding parallel
substitution calculus (CwF-based calculus) are isomorphic. All
CwF-based models are models of our SSC, but not
necessarily the other way. While B-systems are equivalent to CwFs (and
C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
SSC models than CwF-based models. This relationship is similar to that
of extensional combinatory calculus and lambda calculus, where the
syntaxes are equivalent, and every lambda model is a combinatory
model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. Another example is the
relationship between monoids over a set $X$ and nil/cons algebras over
$X$, where every monoid is a nil/cons-algebra, but not the other way;
but it is well-known that free monoids over $X$ (syntax of monoid over
$X$) are the same as lists of $X$ (the syntax of nil/cons algebras
over $X$).

@@

If in addition to having a universe and $\Pi$ types, our type theory
also supports unit and $\Sigma$ types, we can actually derive a
CwF-based model from an SSC-model. The idea is that $\Sigma$ types can
represent contexts, functions between $\Sigma$ types represent
parallel substitutions, functions into the universe represent
dependent types. Then the SSC is just there for bootstrapping
purposes: we use the SSC to formulate $\Sigma$, $\Pi$, $\U$, and then
we use these to define a parallel substitution calculus. In summary,
although SSC models and CwFs without any type formers are not
equivalent, if we equip them with enough type formers, they become
(weakly) equivalent.

All of the constructions in this paper can be understood as happening
internal to extensional type theory, and most results were formalised
in Agda. Some results in this paper were presented at the TYPES 2024
conference \cite{singleTypes}.

@@

\subparagraph*{B-systems.} B-systems introduced by Voevodsky \cite{voevodsky2014bsystems}
are an intrinsic, essentially
algebraic presentation of type theory using single
substitutions. B-systems relate to our SSC as essentially
algebraic theories relate to generalised algebraic theories, or
sets with a map into $I$ relate to indexed families over $I$
\cite[page 221]{DBLP:journals/apal/Cartmell86}. Another difference is
that we have fewer and less general equations, resulting in the fact
that we have more models than B-systems. However in the syntax of our
theory, all the rules of B-systems are admissible. We describe the
relationship in more detail in Appendix \ref{app:bsys}.

Kaposi and Luksa \cite{luksa} defined a telescopic single substitution
calculus for simple type theory, it can be seen as a simply typed
version of B-systems. They show that the category of contextual models
of their calculus is equivalent to the category of contextual simply
typed CwFs with function space.
%Their equivalence result holds even without the presence of function types.
%In the presence of function space,
The same equivalence holds for the simply typed version of our
substitution calculus (TODO link).

@@

\subsection{Metatheory and formalisation}\label{sec:metatheory}

@@

Because we introduced weakening, we now have to explain what happens
to a $\Pi$ type once we weaken it. Assume $A : \Ty\,\Gamma\,i$,
$B:\Ty\,(\Gamma\ext A)\,i$ and we have $\p:\Sub\,(\Gamma\ext
C)\,\Gamma$. Then we say $(\Pi\,A\,B)[\p] = \Pi\,(A[\p])\,(B[\p'])$,
but $B$ cannot be weakened by $\p$ because there we need $\p' :
\Sub\,(\Gamma\ext C\ext A[\p])$ $(\Gamma\ext A)$. So $p'$ has to be a % hack
new kind of weakening which adds a new variable in the last but one
position. But after introducing $\p'$, we would need another equation computing $(\Pi\,A\,B)[\p']$,
introducing a new weakening which adds a new variable in the last but
two position, and so on. We solve all of these issues by allowing the
\emph{lifting} of a weakening: $\gamma^+$ will be the same as the
weakening $\gamma : \Sub\,\Delta\,\Gamma$, except that it will add an
extra variable both at the end of the domain and codomain
context.
\begin{alignat*}{10}
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A)
\end{alignat*}
Note that $A$ is an implicit argument of $\blank^+$, and that it has
to be instantiated by the weakening $\gamma$ in the domain context in
order to fit in. Now we can explain how instantiation acts on $\Pi$ by
the following equation.
\begin{alignat*}{10}
& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+])
\end{alignat*}
The above equation has 6 implicit arguments, namely $\Gamma$, $i$,
$A$, $B$, $\Delta$ and $\gamma$.  This rule works for any weakening,
no matter how many $\blank^+$s have been applied to it. As of now, all
elements of $\Sub$ have the form
\[
\p^{+^n} : \Sub\,(\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}])\,(\Gamma\ext B_1\ext B_2\ext\dots\ext B_n).
\]
where $+^n$ means the $n$-times iteration of $\blank^+$.

@@

Now that we have $[\p][^+]$ for types, we can derive the instantiation
rule for nondependent function space as follows.
\[
(A\Ra B)[\gamma] =
\Pi\,A\,(B[\p])[\gamma] \overset{\Pi[]}{=}
\Pi\,(A[\gamma])\,(B[\p][\gamma^+]) \overset{[\p][^+]}{=}
\Pi\,(A[\gamma])\,(B[\gamma][\p]) =
A[\gamma]\Ra(B[\gamma])
\]

@@

\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
This concludes the definition of our basic type theory. For
reference, we list the same rules again in Appendix \ref{app:tt}.
\end{definition}
The theory can be extended with new type and term formers in the same
fashion: each operation has to be indexed over arbitrary contexts and come
equipped with a substitution rule. There is no need to add more
structural (substitution calculus) rules.\footnote{This is made formal in \cite{DBLP:conf/fscd/KaposiX24} where
these equations are enough to describe the first-order theory of any SOGAT.}
For example, we show how to
extend Definition \ref{def:tt} with $\top$ and $\Sigma$ types.
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
  We extend Definition \ref{def:tt} with the following rules.
  \begin{alignat*}{10}
    & \rlap{$
  \top : \Ty\,\Gamma\,0 \hspace{3em}
  \top[] : \top[\gamma] = \top\hspace{3em}
  \top\eta : t = \tt \hspace{3em}
  \tt : \Tm\,\Gamma\,\top \hspace{3em}
  \tt[] : \tt[\gamma] = \tt
      $} \\
    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong(a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : \blank{,}\blank \\
    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
  \end{alignat*}
  To save space, we compressed the introduction, elimination, $\beta$
  and $\eta$ rules of $\Sigma$ into an isomorphism, and did not list
  substitution law for $\fst$, $\snd$ as they are derivable (see
  below).
\end{definition}

@@

The polymorphic identity function for types at the level 0 is defined
as
\[
\lam\,(\lam\,\q) : \Tm\,\diamond\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)\Big).
\]
Note that readability of this term essentially relies on implicit
arguments. For example, $\q$ takes $(\diamond\ext\U\,0)$ as its first,
$1$ as its second and $\Lift\,(\El\,\q)$ as its third implicit
argument, and these arguments themselves are terms written using
implicit arguments. Inrinsically typed terms are the same as
derivation trees, we illustrate this by deriving this term as follows.
\[
\infer[\text{(*)}]{\lam\,(\lam\,\q) : \Tm\,\diamond\,\big(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))\big)}{\infer{\lam\,\q : \Tm\,(\diamond\ext\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))}{\infer{\q : \Tm\,(\diamond\ext\U\,0\ext\Lift\,(\El\,\q))\,(\Lift\,(\El\,\q)[\p])}{}}}
\]
Note that in step (*), $\lam$ made it sure that $\U\,0$ and
$\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)$ are types at the same level,
which is why we had to lift $\El\,\q$. Another subtlety is
happening when deriving the third implicit argument of $\q$, because we have
to coerce $\q$ along the equality $\U[]$:
\[
\infer{\Lift\,(\El\,\q) : \Ty\,(\diamond\ext\U\,0)\,1}{\infer{\El\,\q : \Ty\,(\diamond\ext\U\,0)\,0}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,(\U\,0)}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,((\U\,0)[\p])}{} && \infer{\U[] : (\U\,0)[\p] = \U\,0}{}}}}
\]
When working informally (or in extensional type theory), we don't
write the coercion, but Agda requires it.

Given a type $A:\Ty\,\diamond\,0$, and $a : \Tm\,\diamond\,A$, we have
\[
\lam\,(\lam\,\q)\cdot\cd\,A\cdot a \overset{\Pi\beta}{=}
\lam\,\q[\langle\cd\,A\rangle]\cdot a \overset{\lam[]}{=}
\lam\,(\q[\langle\cd\,A\rangle^+])\cdot a\, \overset{[\q][^+]}{=}
\lam\,\q\cdot a \overset{\Pi\beta}{=}
\q[\langle a\rangle] \overset{\q[\langle\rangle]}{=}
a
\]
If we want to specify open inputs $A':\Ty\,(\diamond\ext C)\,0$, and
$a' : \Tm\,(\diamond\ext C)\,A'$, we have to weaken our function to
accept them:
\[
\lam\,(\lam\,\q)[\p] : \Tm\,(\diamond\ext C)\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)[\p]\Big).
\]
But we have
$
\lam\,(\lam\,\q)[\p] \overset{\lam[]\,2\text{x}}{=} \lam\,(\lam\,(\q[\p^{++}])) \overset{\q[^+]}{=} \lam\,(\lam\,\q)
$
(with different implicit arguments for the $\lam$s on the left and right hand side), so we are able to apply $\cd\,A'$
and $a'$ just as before. What we cannot do is to directly weaken a
term in context $\diamond$ to an arbitrary context $\Gamma$, where
$\Gamma$ is a metavariable. If we know that $\Gamma$ has lengh $n$,
then we can do the weakening by
$\blank\underbrace{[\p][\p]\dots[\p]}_{\text{$n$ times}}$. If we work
in the syntax, then we can derive such weakenings (and more) using the
methods of Section \ref{sec:admissible}.

@@

$\Sigma$ types were introduced by a similar isomorphism, and we give
names to the roundtrips eqautions according to our convention introduced in
Section \ref{sec:metatheory}: $\Sigma\beta_1 : \fst\,(a,b) = a$,
$\Sigma\beta_2 : \snd\,(a,b) = b$ and $\Sigma\eta : w =
(\fst\,w,\snd\,w)$. Naturality for $\Sigma$ was only stated in one
direction (${,}[]$), naturalities in the other direction are
derivable:
\begin{alignat*}{10}
  & (\fst\,w)[\gamma] && {=}(\Sigma\beta_1)                             && (\snd\,w)[\gamma] && {=}(\Sigma\beta_2)                             \\
  & \fst\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \hspace{6em} && \snd\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \\
  & \fst\,\big((\fst\,w,\snd\,w)[\gamma]\big) && {=}(\Sigma\eta)        && \snd\,\big((\fst\,w,\snd\,w)[\gamma]\big) && {=}(\Sigma\eta)        \\
  & \fst\,(w[\gamma])                                                && && \snd\,(w[\gamma])                                                   
\end{alignat*}
For $\Pi$ types, we construct a similar isomorphism
$
(\lambda t.t[\p]\cdot\q) : \Tm\,\Gamma\,(\Pi\,A\,B) \cong \Tm\,(\Gamma\ext A)\,B : \lam.
$
The first roundtrip equality is proven as
$\beta : (\lam\,t)[\p]\cdot\q \overset{\lam[]}{=} \lam\,(t[\p^+])\cdot\q \overset{\Pi\beta}{=} t[p^+][\langle\q\rangle] \overset{\text{(\ref{eq:lifted4})}}{=} t$.
The last step is the admissible equation (\ref{eq:lifted4}) from
Section \ref{sec:admissible}. As above with weakening into an
arbitrary context, this equation holds in the syntax, bot not in an
arbitrary model.
%and also if $t$ is not a metavariable, but an operation, e.g. if $t = \q$ or $t = \q[\p]\cdot\q$).
The otherroundtrip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.

>>>>

\title{Type theory with single substitutions}

@@

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck the other equations. SSC provides a simple, minimalistic
alternative to parallel substitution calculi or B-systems for defining
type theory. SSC relates to CwF as extensional combinatory calculus
relates to lambda calculus: there are more models of the former, but
the syntaxes are equivalent. If we have some additional type formers,
we show that SSC models are actually equivalent to CwF-based models in
a weak sense.
\end{abstract}

@@

In this paper, we will answer this question by defining type theory as
a generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). GATs are multi-sorted algebraic
theories where later sorts can be indexed by previous sorts. An
example is categories where the sort of morphisms is double-indexed
over objects.
%
The GAT presentation of type theory is intrinsic
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} rather
than extrinsic
\cite{alti:phd93,abel2013normalization,DBLP:conf/cpp/AdjedjLMPP24}:
this means that we only consider well-formed, well-scoped and
well-typed abstract syntax trees, there are no meaningless terms; in
the GAT presentation, the conversion relation is expressed by
equations of the algebraic theory.

@@

Our SSC is in some sense too minimalistic: for example, the equation
$b[\p^+][\langle\q\rangle] = b$ is not \emph{derivable} in any model,
but it is \emph{admissible} in the syntax. (Here $b : \Tm\,(\Gamma\ext
A)\,B$ is a term which depends on a context $\Gamma$ and an extra
variable of type $A$; $b[\p^+] : \Tm\,(\Gamma\ext A\ext
A[\p])\,(B[\p^+])$ is a version where we weakened \emph{before} the
last variable; in $b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$
we substituted the last variable for the previous one.)  This is
analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In this paper, we show
that the \emph{syntaxes} of SSC and the corresponding parallel
substitution calculus (CwF-based calculus) are isomorphic. All
CwF-based models are models of SSC, but not
necessarily the other way. While B-systems are equivalent to CwFs (and
C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
SSC models than CwF-based models. This relationship is like that
of extensional combinatory calculus and lambda calculus, where the
syntaxes are equivalent, and every lambda model is a combinatory
model, but not the other way
\cite{DBLP:conf/fscd/AltenkirchKSV23}. Another example is the
relationship between monoids over a set $X$ and nil/cons algebras over
$X$: every monoid is a nil/cons-algebra, but not the other way;
it is well-known that free monoids over $X$ (syntax of monoid over
$X$) are the same as $X$-lists (the syntax of nil/cons algebras
over $X$).

@@

If in addition to having a universe and $\Pi$ types, our type theory
also supports unit and $\Sigma$ types, we can actually derive a
CwF-based model from an SSC-model. The idea is that $\Sigma$ types can
represent contexts, functions between $\Sigma$ types represent
parallel substitutions, functions into the universe represent
dependent types. Then the SSC is just there for bootstrapping
purposes: SSC formulates $\Sigma$, $\Pi$, $\U$, and then
we use these to define a parallel substitution calculus. In summary,
although SSC models and CwFs without any type formers are not
equivalent, if we equip them with enough type formers, they become
(weakly) equivalent.

All of the constructions in this paper can be understood as happening
internal to extensional type theory, and most results were formalised
in Agda. %Some results in this paper were presented at the TYPES 2024
%conference \cite{singleTypes}.

@@

\subparagraph*{B-systems.} B-systems introduced by Voevodsky \cite{voevodsky2014bsystems}
are an intrinsic, essentially
algebraic presentation of type theory using single
substitutions. B-systems relate to our SSC as essentially
algebraic theories relate to generalised algebraic theories, or
sets with a map into $I$ relate to indexed families over $I$
\cite[page 221]{DBLP:journals/apal/Cartmell86}. Another difference is
that we have fewer and less general equations, resulting in the fact
that we have more models than B-systems. However in the syntax of our
theory, all the rules of B-systems are admissible. We describe the
relationship in more detail in Appendix \ref{app:bsys}.
%
Kaposi and Luksa \cite{luksa} defined a telescopic single substitution
calculus for simple type theory, it can be seen as a simply typed
version of B-systems. They show that the category of contextual models
of their calculus is equivalent to the category of contextual simply
typed CwFs with function space.
%Their equivalence result holds even without the presence of function types.
%In the presence of function space,
The same equivalence holds for the simply typed version of our
substitution calculus (TODO link).

@@

\vspace{-0.7em}
\subsection{Metatheory and formalisation}\label{sec:metatheory}

@@

Because we introduced weakening, we now have to explain what happens
to a $\Pi$ type once we weaken it. Assume $A : \Ty\,\Gamma\,i$,
$B:\Ty\,(\Gamma\ext A)\,i$ and we have $\p:\Sub\,(\Gamma\ext
C)\,\Gamma$. Then we say $(\Pi\,A\,B)[\p] = \Pi\,(A[\p])\,(B[\p'])$,
but $B$ cannot be weakened by $\p$ because there we need $\p' :
\Sub\,(\Gamma\ext C\ext A[\p])$ $(\Gamma\ext A)$. So $p'$ has to be a % hack
new kind of weakening which adds a new variable in the last but one
position. But after introducing $\p'$, we would need another equation computing $(\Pi\,A\,B)[\p']$,
introducing a new weakening which adds a new variable in the last but
two position, and so on. We solve all of these issues by allowing the
\emph{lifting} of a weakening: $\gamma^+$ will be the same as the
weakening $\gamma : \Sub\,\Delta\,\Gamma$, except that it will add an
extra variable both at the end of the domain and codomain
context.
\begin{alignat*}{10}
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A)
\end{alignat*}
Note that $A$ is an implicit argument of $\blank^+$, and that it has
to be instantiated by the weakening $\gamma$ in the domain context in
order to fit in. Now we can explain how instantiation acts on $\Pi$ by
the following equation.
\begin{alignat*}{10}
& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+])
\end{alignat*}
The above equation has 6 implicit arguments, namely $\Gamma$, $i$,
$A$, $B$, $\Delta$ and $\gamma$.  This rule works for any weakening,
no matter how many $\blank^+$s have been applied to it. As of now, all
elements of $\Sub$ have the form
$
\p^{+^n} : \Sub\,(\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}])\,(\Gamma\ext B_1\ext B_2\ext\dots\ext B_n),
$
where $^{+^n}$ means the $n$-times iteration of $\blank^+$.

@@

Now that we have $[\p][^+]$ for types, we can derive its instantiation rule by
$
(A\Ra B)[\gamma] =
\Pi\,A\,(B[\p])[\gamma] \overset{\Pi[]}{=}
\Pi\,(A[\gamma])\,(B[\p][\gamma^+]) \overset{[\p][^+]}{=}
\Pi\,(A[\gamma])\,(B[\gamma][\p]) =
(A[\gamma])\Ra(B[\gamma]).
$

@@

\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
This concludes the definition of our basic type theory. For
reference, we list the same rules again in Appendix \ref{app:tt}.
\end{definition}
The theory can be extended with new type and term formers in the same
fashion: each operation has to be indexed over arbitrary contexts and come
equipped with a substitution rule. There is no need to add more
structural (substitution calculus) rules.\footnote{In \cite{DBLP:conf/fscd/KaposiX24},
the same structural rules suffice to describe the first-order theory for any SOGAT.}
For example, we show how to
extend Definition \ref{def:tt} with $\top$ and $\Sigma$ types.
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
  We extend Definition \ref{def:tt} with the following rules.
  \begin{alignat*}{10}
    & \rlap{$
  \top : \Ty\,\Gamma\,0 \hspace{3em}
  \top[] : \top[\gamma] = \top\hspace{3em}
  \top\eta : t = \tt \hspace{3em}
  \tt : \Tm\,\Gamma\,\top \hspace{3em}
  \tt[] : \tt[\gamma] = \tt
      $} \\
    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong(a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : \blank{,}\blank \\
    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
  \end{alignat*}
  To save space, we compressed the introduction, elimination, $\beta$
  and $\eta$ rules of $\Sigma$ into an isomorphism, and did not list
  substitution law for $\fst$, $\snd$ as they are derivable (see
  below).
\end{definition}

@@

The polymorphic identity function for types at the level 0 is defined
as
\[
\lam\,(\lam\,\q) : \Tm\,\diamond\,(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))).
\]
Note that readability of this term essentially relies on implicit
arguments. For example, $\q$ takes $(\diamond\ext\U\,0)$ as its first,
$1$ as its second and $\Lift\,(\El\,\q)$ as its third implicit
argument, and these arguments themselves are terms written using
implicit arguments. Intrinsically typed terms are the same as
derivation trees, we illustrate this by deriving this term as follows.
\[
\infer[\text{(*)}]{\lam\,(\lam\,\q) : \Tm\,\diamond\,\big(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))\big)}{\infer{\lam\,\q : \Tm\,(\diamond\ext\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))}{\infer{\q : \Tm\,(\diamond\ext\U\,0\ext\Lift\,(\El\,\q))\,(\Lift\,(\El\,\q)[\p])}{}}}
\]
Note that in step (*), $\lam$ made it sure that $\U\,0$ and
$\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)$ are types at the same level,
which is why we had to lift $\El\,\q$. Another subtlety is
happening when deriving the third implicit argument of $\q$, because we have
to coerce $\q$ along the equality $\U[]$:
\[
\infer{\Lift\,(\El\,\q) : \Ty\,(\diamond\ext\U\,0)\,1}{\infer{\El\,\q : \Ty\,(\diamond\ext\U\,0)\,0}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,(\U\,0)}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,((\U\,0)[\p])}{} && \infer{\U[] : (\U\,0)[\p] = \U\,0}{}}}}
\]
When working informally (or in extensional type theory), we don't
write coercions.%, but Agda requires it.

Given a type $A:\Ty\,\diamond\,0$, and $a : \Tm\,\diamond\,A$, we have
$
\lam\,(\lam\,\q)\cdot\cd\,A\cdot a \overset{\Pi\beta}{=}
\lam\,\q[\langle\cd\,A\rangle]\cdot a \overset{\lam[]}{=}
\lam\,(\q[\langle\cd\,A\rangle^+])\cdot a\, \overset{[\q][^+]}{=}
\lam\,\q\cdot a \overset{\Pi\beta}{=}
\q[\langle a\rangle] \overset{\q[\langle\rangle]}{=}
a.
$
If we want to specify open inputs $A':\Ty\,(\diamond\ext C)\,0$, and
$a' : \Tm\,(\diamond\ext C)\,A'$, we have to weaken our function to
accept them:
$
\lam\,(\lam\,\q)[\p] : \Tm\,(\diamond\ext C)\,(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))[\p]).
$
But we have
$
\lam\,(\lam\,\q)[\p] \overset{\lam[]\,2\text{x}}{=} \lam\,(\lam\,(\q[\p^{++}])) \overset{\q[^+]}{=} \lam\,(\lam\,\q)
$
(with different implicit arguments for the $\lam$s on the left and right hand side), so we are able to apply $\cd\,A'$
and $a'$ just as before. What we cannot do is to directly weaken a
term in context $\diamond$ to an arbitrary context $\Gamma$, where
$\Gamma$ is a metavariable. If we know that $\Gamma$ has length $n$,
then we can do the weakening by applying $\blank[\p]$ to it $n$ times.
If we work
in the syntax, then we can derive such weakenings (and more) using the
methods of Section \ref{sec:admissible}.

@@

$\Sigma$ types were introduced by a similar isomorphism, and we give
names to the roundtrips equations according to our convention introduced in
Section \ref{sec:metatheory}: $\Sigma\beta_1 : \fst\,(a,b) = a$,
$\Sigma\beta_2 : \snd\,(a,b) = b$ and $\Sigma\eta : w =
(\fst\,w,\snd\,w)$. Naturality for $\Sigma$ was only stated in one
direction (${,}[]$), naturalities in the other direction are
derivable, for $\fst$ this is
$
(\fst\,w)[\gamma]  \overset{\Sigma\beta_1}{=}                          
\fst\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big) \overset{{,}[]}{=} 
\fst\,\big((\fst\,w,\snd\,w)[\gamma]\big)  \overset{\Sigma\eta}=        
\fst\,(w[\gamma]).
$
For $\Pi$ types, we construct a similar isomorphism
$
(\lambda t.t[\p]\cdot\q) : \Tm\,\Gamma\,(\Pi\,A\,B) \cong \Tm\,(\Gamma\ext A)\,B : \lam.
$
The first roundtrip equality is proven as
$\beta : (\lam\,t)[\p]\cdot\q \overset{\lam[]}{=} \lam\,(t[\p^+])\cdot\q \overset{\Pi\beta}{=} t[p^+][\langle\q\rangle] \overset{\text{(\ref{eq:lifted4})}}{=} t$.
The last step is the admissible equation (\ref{eq:lifted4}) from
Section \ref{sec:admissible}. As above with weakening into an
arbitrary context, this equation holds in the syntax, bot not in an
arbitrary model.
%and also if $t$ is not a metavariable, but an operation, e.g. if $t = \q$ or $t = \q[\p]\cdot\q$).
The other roundtrip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.

We construct another isomorphism between terms depending on a variable and
terms depending on the lifted version of the same variable:
\begin{equation}\label{eq:liftvar}
(\lambda t.t[\p^+][\langle\un\,\q\rangle]) : \Tm\,(\Gamma\ext A)\,B \cong \Tm\,(\Gamma\ext\Lift\,A)\,(B[\p^+][\langle\q\rangle]) : (\lambda t.t[\p^+][\langle\mk\,\q\rangle])
\end{equation}
To derive the roundtrips, we again need admissible equations. We prove
the first roundtrip equality in Appendix \ref{app:liftvar}, the other
is analogous.
Once we have all the above isomorphisms, we can compose them to obtain commutation of $\Lift$ and $\Pi$:
\begin{alignat*}{10}
  & \Tm\,\Gamma\,\big(\Lift\,(\Pi\,A\,B)\big) \overset{\text{$\Lift$}}{\cong}
  \Tm\,\Gamma\,(\Pi\,A\,B) \overset{\text{$\Pi$}}{\cong} 
  \Tm\,(\Gamma\ext A)\,B \overset{\text{$\Lift$}}{\cong} 
  \Tm\,(\Gamma\ext A)\,(\Lift\,B) \overset{\text{lifted var}}{\cong} \\
  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) \overset{\text{$\Pi$}}{\cong}
  \Tm\,\Gamma\,\big(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle])\big)
\end{alignat*}
We hope that the above examples demonstrated the need for equations
(\ref{eq:lifted1})--(\ref{eq:lifted4}) which we will prove in the next
section.

% \subsection{Standard model}
% \label{sec:standard}
% 
% As a sanity check for our notion of SSC-model, we defined the standard
% (metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model of the
% SSC-based calculus in Agda. For convenience, we used an
% inductive-recursive universe to define the infinite hierarchy of types
% \cite{DBLP:journals/jsyml/Dybjer00,DBLP:conf/csl/Kovacs22}. All
% SSC-equations hold by reflexivity in this model. We defined the
% analogous CwF-based standard model.
